A full approximation scheme multilevel method \\ for nonlinear variational inequalities

### 2023-12-24 paper/fascd.tex
edits and clean up through most of section 2

The paper is organized as follows.  In Section \ref{sec:vi} we recall the theory of coercive and monotone VIs, and describe several motivating examples.  Section \ref{sec:cd} reviews CDs and their induced iterations.  Section \ref{sec:femultilevel} sets up multilevel FE hierarchies, and then Section \ref{sec:cdmultilevel} implements the CD method using defect constraints which are generated by monotone injection \cite{GraeserKornhuber2009}.  In Section \ref{sec:cdmultilevel} we also propose a new understanding of telescoping sets for up-smoothing, based on an ``incomplete'' CD and an associated iteration.  Section \ref{sec:vcycle} then states the FASCD V-cycle, and Section \ref{sec:implementation} addresses practical implementation including convergence criteria and the FMG cycle.  Section \ref{sec:results} gives numerical results from our open source Python/Firedrake \cite{Rathgeberetal2016} implementation.

@@

Let $\cV$ be a real reflexive Banach space with norm $\|\cdot\|$ and topological dual space $\cV'$.  Denote the dual pairing of $\phi \in \cV'$ and $v\in\cV$ by $\ip{\phi}{v} = \phi(v)$, and define the (Banach space) norm on $\cV'$ by $\|\phi\|' = \sup_{\|v\|=1} |\!\ip{\phi}{v}\!|$.  Let $\cK \subset \cV$ be a nonempty, closed, and convex subset, the \emph{constraint set}; elements of $\cK$ are said to be \emph{admissible}.  For a continuous, but generally nonlinear, operator $f:\cK \to \cV'$ and a linear \emph{source functional} $\ell\in \cV'$ we consider the following \emph{variational inequality} (VI) problem: find $u^\star\in \cK$ such that
\begin{equation}
\ip{f(u^\star)}{v-u^\star} \ge \ip{\ell}{v-u^\star} \quad \text{for all } v\in \cK. \label{eq:vi}
\end{equation}
Because $f$ is (generally) nonlinear, the source term $\ell$ is not strictly needed in stating this class of problems---by redefining $f$ we may take $\ell=0$---but introducing $\ell$ clarifies the algorithms which follow.

VI \eqref{eq:vi} generalizes nonlinear systems of equations $f(u^\star)=\ell$ to problems where $u^\star$ is also constrained to be in $\cK$.  Pretending the dual pairing is an inner product, by \eqref{eq:vi} the angle between $f(u^\star)-\ell$ and any arbitrary vector $v-u^\star$ pointing from $u^\star$ into $\cK$ is at most $90^\circ$; the dual vector $f(u^\star)-\ell$ must point into $\cK$.  If $u^\star$ is in the interior of the constraint set then \eqref{eq:vi} implies $f(u^\star)=\ell$.
Variational inequalities may generally permit multiple solutions~\cite{Farrell2019}.

The following definitions are standard \cite{KinderlehrerStampacchia1980}.

@@

It is well-known that if $f:\cK \to \cV'$ is continuous, monotone, and coercive then VI \eqref{eq:vi} has a solution \cite[Corollary III.1.8]{KinderlehrerStampacchia1980}, and that the solution is unique when $f$ is strictly monotone.  As is standard in the calculus of variations \cite{Evans2010}, coercivity permits a compactness argument for unbounded sets $\cK$; recall that the closed and bounded subsets of a reflexive Banach space are weakly compact.  Some VIs solved in this paper satisfy a stronger inequality than \eqref{eq:coercive}.

\begin{definition}  Let $p>1$.  The map $f:\cK \to \cV'$ is \emph{$p$-coercive} if there exists $\kappa>0$ such that
\begin{equation}
\ip{f(u)-f(v)}{u-v} \ge \kappa \|u-v\|^p \qquad \text{for all } u,v \in \cK. \label{eq:pcoercive}
\end{equation}
\end{definition}

It is easy to see that if $f$ is $p$-coercive then it is monotone, strictly monotone, and coercive.  Thus if $f:\cK \to \cV'$ is continuous and $p$-coercive then there exists a unique $u^\star\in \cK$ solving VI \eqref{eq:vi}.

Let $\Omega \subset \RR^d$ denote a bounded, open set with piecewise-smooth boundary.  Sobolev spaces \cite{Evans2010} will be denoted by $W^{k,p}(\Omega)$, for integer $k$ and $1\le p \le \infty$, with $W^{1,p}_0(\Omega)$ denoting the subspace with zero trace.  The following example includes the classical and $p$-Laplacian obstacle problems \cite{ChoeLewis1991}.

\begin{example}  \label{ex:plaplacian}  Let $1<p<\infty$.  For $u,v \in \cV = W^{1,p}_0(\Omega)$, define $f:\cV \to \cV'$ by
\begin{equation}
\ip{f(u)}{v} = \int_\Omega |\grad u|^{p-2} \grad u \cdot \grad v\dx, \label{eq:plaplacian}
\end{equation}
a continuous map \cite[Theorem A.0.6]{Peral1997}.  It follows from inequalities for $|\cdot|^p$ in $\RR^d$, and the Poincar\'e inequality, that $f$ is $p$-coercive \cite{ChoeLewis1991}; \cite[Appendix A]{Bueler2021conservation}.  \end{example}

For monotone operators $f$, VI \eqref{eq:vi} generalizes the problem of minimizing a convex function over $\cK$.  Suppose $F:\cK \to \RR$ is lower semi-continuous and (G\^ateaux) differentiable with continuous derivative $F':\cK \to \cV'$.  Then $F$ is convex if and only if $F'$ is monotone \cite[Proposition I.5.5]{EkelandTemam1976}.  Furthermore, Proposition II.2.1 in \cite{EkelandTemam1976} shows that if $F$ is convex then \eqref{eq:vi} holds for $f=F'$ if and only if
\begin{equation}
u^\star = \argmin_{v\in\cK} F(v) - \ip{\ell}{v}. \label{eq:minimization}
\end{equation}
The CD methods of Tai \cite{Tai2003} address problem \eqref{eq:minimization}; the analysis in \cite{Tai2003} supposes that an objective $F$ exists with $F'$ 2-coercive.  For Example \ref{ex:plaplacian} we may define
\begin{equation}
F(v) = \frac{1}{p} \int_\Omega |\grad v|^p\dx. \label{eq:plaplacianobjective}
\end{equation}
Then $F$ is a convex functional and $F'(v) = f(v)$ for $f$ given in \eqref{eq:plaplacian}.  For any closed and convex $\cK\subset \cV$, VI problem \eqref{eq:vi} is then equivalent to optimization problem \eqref{eq:minimization}.

@@

\begin{example}  \label{ex:advectiondiffusion}  Suppose $\partial\Omega$ is partitioned into Dirichlet and Neumann portions, $\partial\Omega = \partial_D\Omega \cup \partial_N\Omega$, with $\partial_D\Omega$ of positive Hausdorff measure.  Let $\cV \subset W^{1,2}(\Omega)$ be the space of functions which have zero trace on $\partial_D\Omega$.  Consider a velocity field $\bX$ on $\Omega$ satisfying the conditions of Lemma \ref{lem:advectionskew}, and assume that the flow is outward on the Neumann boundary: $\bX \cdot \bn \ge 0$ on $\partial_N\Omega$.  For $u,v \in \cV$, $\eps>0$, and $\phi \in \cV'$, define
\begin{equation}
\ip{f(u)}{v} = \eps \left(\grad u, \grad v\right)_{L^2(\Omega)} + b(u,v) - \ip{\phi}{v}. \label{eq:advectiondiffusion}
\end{equation}
With this operator, consider VI \eqref{eq:vi} with $\ell = 0$ for any closed and convex $\cK \subset \cV$. The interior condition in strong form is the following linear advection-diffusion equation
\begin{equation}
-\eps \grad^2 u + \bX\cdot \grad u = \phi.
\label{eq:advectiondiffusionstrong}
\end{equation}
On the other hand, it is easy to see that $|\!\ip{f(u)}{v}\!| \le \left((\eps + \|\bX\|_\infty) \|u\| + \|\phi\|'\right) \|v\|$, so that $f:\cK \to \cV'$ is continuous.  Lemma \ref{lem:advectionskew} says that the bilinear form $b(u,v)$ is skew-symmetric up to a term which is nonnegative by the outward flow assumption.  From the Poincar\'e inequality,
\begin{align*}
\ip{f(u)-f(v)}{u-v} &= \eps \int_\Omega |\grad u - \grad v|^2\dx + b(u-v,u-v) \\
                    &= \eps \int_\Omega |\grad u - \grad v|^2\dx + \frac{1}{2} \int_{\partial_N\Omega} (u-v)^2 \bX\cdot\bn \ge \eps C \|u-v\|^2.
\end{align*}
Thus $f$ is 2-coercive, and so the associated VI problem is well-posed. If $\bX \ne 0$ then $f\ne F'$ for any objective $F$, because $\ip{f(u)}{v}$ does not possess the symmetry of a gradient: for $u,v$ which are zero on $\partial \Omega$, note that $\ip{f(u)}{v} - \ip{f(v)}{u} = -2 b(u,v)$.  References \cite{Bueler2021conservation,ChangNakshatrala2017} consider similar advection-diffusion VI problems over $\cK = \{v\ge 0\}$. \end{example}

The next VI example is a model for steady ice sheet geometry in a given climate.  It is fully nonlinear and not of optimization type.  Related VI problems arise for any fluid layer subject to processes which can remove fluid mass \cite{Bueler2021conservation}.

>>>>

The paper is organized as follows.  In Section \ref{sec:vi} we recall the theory of coercive and monotone VIs, and describe several motivating examples.  Section \ref{sec:cd} reviews CDs and their induced iterations.  Section \ref{sec:femultilevel} sets up multilevel FE hierarchies, and then Section \ref{sec:cdmultilevel} implements the CD method using defect constraints which are generated by monotone injection.  In Section \ref{sec:cdmultilevel} we also propose a new understanding of telescoping sets for up-smoothing, based on an ``incomplete'' CD and an associated iteration.  Section \ref{sec:vcycle} then states the FASCD V-cycle, and Section \ref{sec:implementation} addresses practical implementation including convergence criteria and the FMG cycle.  Section \ref{sec:results} gives numerical results from our open source Python/Firedrake \cite{Rathgeberetal2016} implementation.

@@

Let $\cV$ be a real reflexive Banach space with norm $\|\cdot\|$ and topological dual space $\cV'$.  Denote the dual pairing of $\phi \in \cV'$ and $v\in\cV$ by $\ip{\phi}{v} = \phi(v)$, and define the (Banach space) norm on $\cV'$ by $\|\phi\|' = \sup_{\|v\|=1} |\!\ip{\phi}{v}\!|$.  Let $\cK \subset \cV$ be a nonempty, closed, and convex subset, the \emph{constraint set}; elements of $\cK$ are said to be \emph{admissible}.  For a continuous, but generally nonlinear, operator $f:\cK \to \cV'$ and a linear \emph{source functional} $\ell\in \cV'$ we consider the following \emph{variational inequality} (VI) problem: find $u^\star\in \cK$ such that
\begin{equation}
\ip{f(u^\star)}{v-u^\star} \ge \ip{\ell}{v-u^\star} \quad \text{for all } v\in \cK. \label{eq:vi}
\end{equation}
Because $f$ is (generally) nonlinear, the source term $\ell$ is not strictly needed in stating this class of problems---by redefining $f$ we may take $\ell=0$---but introducing $\ell$ will help to clarify later algorithms.

VI \eqref{eq:vi} generalizes nonlinear systems of equations (i.e.~$f(u^\star)=\ell$) to cases where $u^\star$ is also constrained to be in $\cK$.  Pretending the dual pairing is an inner product, \eqref{eq:vi} says that the angle between $f(u^\star)-\ell$ and any arbitrary vector $v-u^\star$ pointing from $u^\star$ into $\cK$ is at most $90^\circ$; if $u^\star$ solves \eqref{eq:vi} then the dual vector $f(u^\star)-\ell$ points straight into $\cK$.  If $u^\star$ is in the interior of $\cK$ then \eqref{eq:vi} implies $f(u^\star)=\ell$.
Note that variational inequalities generally permit multiple solutions~\cite{Farrell2019}.

The following definitions are standard \cite[for example]{KinderlehrerStampacchia1980}.

@@

It is well-known that if $f:\cK \to \cV'$ is continuous, monotone, and coercive then VI \eqref{eq:vi} has a solution \cite[Corollary III.1.8]{KinderlehrerStampacchia1980}, and that the solution is unique when $f$ is strictly monotone.  As in the calculus of variations, coercivity permits a compactness argument for unbounded sets $\cK$ using the fact that closed and bounded subsets of a reflexive Banach space are weakly compact.

Some operators considered below satisfy a stronger inequality than \eqref{eq:coercive}.

\begin{definition}  Let $p>1$.  The operator $f:\cK \to \cV'$ is \emph{$p$-coercive} if there exists $\kappa>0$ such that
\begin{equation}
\ip{f(u)-f(v)}{u-v} \ge \kappa \|u-v\|^p \qquad \text{for all } u,v \in \cK. \label{eq:pcoercive}
\end{equation}
\end{definition}

It is easy to see that if $f$ is $p$-coercive then it is monotone, strictly monotone, and coercive.  Thus if $f:\cK \to \cV'$ is continuous and $p$-coercive ($p>1$) then there exists a unique $u^\star\in \cK$ solving VI \eqref{eq:vi}.

Let $\Omega \subset \RR^d$ denote a bounded, open set with piecewise-smooth boundary.  Sobolev spaces \cite{Evans2010} will be denoted by $W^{k,p}(\Omega)$, for integer $k$ and $1\le p \le \infty$, with $W^{1,p}_0(\Omega)$ denoting the subspace with zero trace along $\partial\Omega$.  The following example includes the classical and $p$-Laplacian obstacle problems \cite{ChoeLewis1991}.

\begin{example}  \label{ex:plaplacian}  Let $1<p<\infty$.  For $u,v \in \cV = W^{1,p}_0(\Omega)$, define $f:\cV \to \cV'$ by
\begin{equation}
\ip{f(u)}{v} = \int_\Omega |\grad u|^{p-2} \grad u \cdot \grad v\dx. \label{eq:plaplacian}
\end{equation}
It follows from inequalities for $|\cdot|^p$ in $\RR^d$, and the Poincar\'e inequality, that $f$ is $p$-coercive \cite{ChoeLewis1991}; see also \cite[Appendix A]{Bueler2021conservation}.  \end{example}

Note that for monotone operators VI \eqref{eq:vi} also generalizes the problem of minimizing a convex function over $\cK$.  Suppose $F:\cK \to \RR$ is lower semi-continuous and (G\^ateaux) differentiable with continuous derivative $F':\cK \to \cV'$.  Then $F$ is convex if and only if $f=F'$ is monotone, in which case \eqref{eq:vi} holds if and only if
\begin{equation}
u^\star = \argmin_{v\in\cK} F(v) - \ip{\ell}{v}. \label{eq:minimization}
\end{equation}
(See e.g.~Propositions I.5.5 and II.2.1 in \cite{EkelandTemam1976} for these facts.)  The CD methods of Tai \cite{Tai2003} address problem \eqref{eq:minimization}; his analysis supposes that an objective $F$ exists with $F'$ 2-coercive.  In Example \ref{ex:plaplacian}
\begin{equation}
F(v) = \frac{1}{p} \int_\Omega |\grad v|^p\dx \label{eq:plaplacianobjective}
\end{equation}
is a convex functional and $F' = f$ is given by \eqref{eq:plaplacian}.

@@

\begin{example}  \label{ex:advectiondiffusion}  Suppose $\partial\Omega$ is partitioned into Dirichlet and Neumann portions, $\partial\Omega = \partial_D\Omega \cup \partial_N\Omega$, with $\partial_D\Omega$ of positive Hausdorff measure.  Let $\cV \subset W^{1,2}(\Omega)$ be the space of functions which have zero trace on $\partial_D\Omega$.  Consider a velocity field $\bX$ on $\Omega$ satisfying the conditions of Lemma \ref{lem:advectionskew}, and also assume that the flow is outward on the Neumann boundary: $\bX \cdot \bn \ge 0$ on $\partial_N\Omega$.  For $u,v \in \cV$, $\eps>0$, and $\phi \in \cV'$, define
\begin{equation}
\ip{f(u)}{v} = \eps \left(\grad u, \grad v\right)_{L^2(\Omega)} + b(u,v) - \ip{\phi}{v}. \label{eq:advectiondiffusion}
\end{equation}
With this operator, consider VI \eqref{eq:vi} with $\ell = 0$ for any closed and convex $\cK \subset \cV$. The interior condition \cite{KinderlehrerStampacchia1980} of \eqref{eq:vi} is the following linear advection-diffusion equation:
\begin{equation}
-\eps \grad^2 u + \bX\cdot \grad u = \phi.
\label{eq:advectiondiffusionstrong}
\end{equation}
On the other hand, it is easy to see that $|\!\ip{f(u)}{v}\!| \le \left((\eps + \|\bX\|_\infty) \|u\| + \|\phi\|'\right) \|v\|$, so $f:\cK \to \cV'$ is continuous.  Lemma \ref{lem:advectionskew} says that the bilinear form $b(u,v)$ is skew-symmetric up to a term which is nonnegative by the outward flow assumption.  From the Poincar\'e inequality,
\begin{align*}
\ip{f(u)-f(v)}{u-v} &= \eps \int_\Omega |\grad u - \grad v|^2\dx + b(u-v,u-v) \\
                    &= \eps \int_\Omega |\grad u - \grad v|^2\dx + \frac{1}{2} \int_{\partial_N\Omega} (u-v)^2 \bX\cdot\bn \ge \eps C \|u-v\|^2.
\end{align*}
Thus $f$ is 2-coercive and the associated VI problem is well-posed. If $\bX \ne 0$ then $f\ne F'$ for any objective $F$, because $\ip{f(u)}{v}$ does not possess the symmetry of a gradient: for $u,v$ which are zero on $\partial \Omega$, note that $\ip{f(u)}{v} - \ip{f(v)}{u} = -2 b(u,v)$.\end{example}

References \cite{Bueler2021conservation,ChangNakshatrala2017} consider similar advection-diffusion VI problems to Example \ref{ex:advectiondiffusion}, over $\cK = \{v\ge 0\}$.

The next VI example is a model for steady ice sheet geometry in a given climate.  It is fully nonlinear and not of optimization type.  Related VI problems arise for any fluid layer subject to local surface processes which can remove fluid mass \cite{Bueler2021conservation}.
